Early Stage Assessment

### 2018-12-03 papers/esa/esa.tex
Edits to ESA

This document is to provide a brief description of my research area which will be the topic of my thesis. All code used thus far and from hence forth will mainly be written in the Julia programming language.

@@

We can the define the spherical harmonics, orthogonal on the unit sphere as \cite{DLMFSphericalandSpheroidalHarmonics, DLMFFerrers, DLMFAssociatedLegendre}:
\begin{align}
\Ylmfull &:= \clm \eimphi \Plm (\costheta) \\
&= \alphalm (1 - (\costheta)^2)^\frac{|m|}{2} \eimphi P^{(|m|,|m|)}_{l-|m|}(\costheta), \quad \text{where } 0 \le |m| \le l, \, l \in \No.
\end{align} 
Note that we can then express \(\Ylm\) in terms of \(x,y,z\) instead of \(\theta, \varphi\) by noting that \(\costheta = z\) and that \( \eimphi\) can be expressed in terms of \(x,y,z\) for any \(m\in\mathbb{Z}\). Indeed, they are polynomials in $x,y,z$ which we denote $\Ylm(x,y,z)$. They span all polynomials modulo the ideal generated by $x^2 + y^2 + z^2 - 1$.

\citep{dunkl2014orthogonal} provided a framework for gaining a recurrence relation for each \(\bigP_{l+1}\) in terms of the previous two sub-vectors \(\Pl\) and \(\bigP_{l-1}\), utilising the sub matrices that make up the block-tridiagnonal Jacobi operator matrices for multiplication by $x, y$ and $z$. Clenshaw's algorithm then gives us a way of evaluating a function given its expansion in the spherical harmonic basis (up to a finite order).

@@

For \(l \in \No\), \(m \in \Z\) s.t. \(0 \le |m| \le l\):
\begin{align}
x\,\Ylm(x,y,z) &= \Alm Y^{m+1}_{l+1}(x,y,z) +  \Blm Y^{m+1}_{l-1}(x,y,z) \nonumber \\
		     & \quad \quad \quad + \Dlm Y^{m-1}_{l+1}(x,y,z) + \Elm Y^{m-1}_{l-1}(x,y,z), \\
y\,\Ylm(x,y,z) &= - i \, \big[\Alm Y^{m+1}_{l+1}(x,y,z) +  \Blm Y^{m+1}_{l-1}(x,y,z) \big] \nonumber \\
		     &\quad \quad \quad + i \, \big[ \Dlm Y^{m-1}_{l+1}(x,y,z) + \Elm Y^{m-1}_{l-1}(x,y,z) \big], \\
z\,\Ylm(x,y,z) &= \Flm Y^{m}_{l+1}(x,y,z) + \Glm Y^{m}_{l-1}(x,y,z) ,
\end{align}
where \(\Alm, \Blm, \Dlm, \Elm, \Flm, \Glm \in \R\) have derivable and calculable formulae.

@@

For each \(l \in \No\) let \(\Dlt\) be any matrix that is a left inverse of \(A_l\), i.e. such that \(\Dlt A_l = I_{2l+3}\). Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the \(\Dlt\)'s, we obtain a lower triangular system \citep{dunkl2014orthogonal}, which can be expanded to obtain the recurrence:
\begin{align}
\begin{cases}
\bigP_{-1} := 0 \\
\bigP_{0} := \alpha^0_0 \\
\bigP_{l+1} = -\Dlt (B_l-G_l) \Pl - \Dlt C_l  \,\bigP_{l-1}, \quad l \in \N.
\end{cases}
\end{align}

\subsection{Evaluation of a scalar function on the sphere}
We can use the Clenshaw algorithm to evaluate a function at a given point \((x,y,z)\) on the unit sphere provided we know the coefficients of the function when expanded in the spherical harmonic basis, i.e. suppose \(f(x,y,z)\) is a function and we know the set \(\{\bold{f}_l\}\) s.t.
\begin{align}
f(x,y,z) \approx \sum^N_{l=0} \bold{f}_l^T \, \Pl (x,y,z), \quad \text{where } \Pl (x,y,z), \bold{f}_l \in \R^{2l+1} \text{ for each } l \in \{0,\dots,N\}.
\end{align}

@@

By noting that the tangent space is then spanned by $\{\gradYlm, \gradpYlm\}$, and that these gradients and perpendicular gradients are orthogonal polynomials themselves, we can use the same ideas to create a framework for functions in the tangent space as expansions in this OP basis.

Since the spherical harmonics are a basis for the surface of the sphere, and the tangent space of the sphere is spanned by the gradient  and perpendicular gradient of a scalar function, we have that the gradients and perpendicular gradients of the spherical harmonics are a basis for the tangent space, namely \(\gradYlm\), \(\gradpYlm\). Note that the perpendicular gradient is related to the regular surface gradient by
\begin{align}
\gradpYlm(x,y,z) = \unitvec \times \gradYlm(x,y,z),
\end{align}
where \(\unitvec\) is the unit vector normal to the surface of the sphere at the point \((x,y,z)\), i.e. as we are looking at the unit sphere, \(\unitvec\) is simply given by
\begin{align}
\unitvec = \begin{bmatrix} x \\ y \\ z \end{bmatrix}.
\end{align}

@@

The gradient and perpendicular gradient of a spherical harmonic \(\Ylm(x,y,z)\) can be expressed in terms of spin-1 tensor spherical harmonics, which in turn can each be expressed as a vector-weighted sum of spherical harmonics. For the derivation and more on vector spherical harmonics, see the appendix.

@@

Then using the operators on the coefficient vectors, the timestepping method is: 
\begin{align}
\mathbf{u}^c_{n+1} &= \mathbf{u}^c_{n} + \Delta t \: (G \mathbf{h}^c_{n+1} - F K \mathbf{u}^c_{n+1}) \\
\mathbf{h}^c_{n+1} &= \mathbf{h}^c_n - \Delta t \: \mathcal{H} D \mathbf{u}^c_{n+1}
\end{align}
$\iff$
\begin{align}
\mathbf{u}^c_{n+1} &= (I + \Delta t F K + \Delta t^2 \mathcal{H} G D)^{-1} \: (\mathbf{u}^c_{n} + \Delta t \: G \mathbf{h}^c_n) \\
\mathbf{h}^c_{n+1} &= \mathbf{h}^c_n - \Delta t \: \mathcal{H} D \mathbf{u}^c_{n+1}
\end{align}

@@

As a stepping stone to working on spherical triangles, we will try to construct orthogonal polynomials on a circle arc, given by $x \in [h, 1]$, $y = (1-x^2)^\frac{1}{2}$. We will use the lanczos algorithm from ApproxFun \cite{ApproxFun} to obtain 1D OPs $\{T^h_n\}$ (orthogonal wrt $w(x) = (1-x^2)^{-\frac{1}{2}}$) and $\{U^h_n\}$ (orthogonal wrt $w(x) = (1-x^2)^{\frac{1}{2}}$) on the interval $[h,1]$, and create our 2D OPs according to:
\begin{align}
\bigP_0(x,y) := T^h_0(x) \equiv T_0,
\quad \quad 
\bigP_n(x,y) := \begin{bmatrix}
		T^h_n(x) \\
		y \: U^h_{n-1}(x)
	\end{bmatrix} \in \R^2 \quad \forall n \in \N,
\quad \quad 
\bigP := \begin{bmatrix}
		\bigP_0 \\
		\hline
		\bigP_1 \\
		\hline
		\bigP_2 \\
		\vdots \\
	\end{bmatrix}.
\end{align}

@@

By using a similar process to \cite{dunkl2014orthogonal} we can construct 2D orthogonal polynomials on the half disk from 1D orthogonal polynomials on the intervals \([0,1]\) and \([-1,1]\). We denote the half disk by \(\Omega\) where
\begin{align}
\Omega := \{(x,y) \in \R^2 \quad | \quad 0 < x < 1, \: -(1-x^2)^{\half} < y < (1-x^2)^{\half}\}.
\end{align}

@@

Fix \(a,b \in \R\). Then for any function \(f : \Omega \to \R\) we can approximate (or exactly express) \(f\) by
\begin{align}
f(x,y) + \sum_{n=0}^N \bigP_n^{(a,b)}(x,y)^T \: \mathbf{f}_n
\end{align}
where
\begin{align}
\bigP(x,y) := \begin{bmatrix}
		\bigP_0(x,y) \\
		\hline
		\bigP_1(x,y) \\
		\hline
		\bigP_2(x,y) \\
		\hline
		\vdots \\
		\hline
		\bigP_N(x,y)
	\end{bmatrix},
\quad
\bigP_n(x,y) := \begin{bmatrix}
		P_{n,0}(x,y) \\
		\vdots \\
		P_{n,n}(x,y)
	\end{bmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N,
\end{align}
and where
\begin{align}
\mathbf{f} &:= \begin{bmatrix}
		\mathbf{f}_0 \\
		\hline
		\mathbf{f}_1 \\
		\hline
		\mathbf{f}_2\\
		\hline
		\vdots \\
		\hline
		\mathbf{f}_N
	\end{bmatrix},
\quad
\mathbf{f}_n := \begin{bmatrix}
		f_{n,0} \\
		\vdots \\
		f_{n,n}
	\end{bmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N, \\
f_{n,k} &:= \frac{< f, \: \Pnk^{(a,b)} >_{W^{(a,b)}}}{|| \Pnk^{(a,b)} ||_{W^{(a,b)}}}
\end{align}

@@

For each \(n = 0,1,\dots,N-1\) let \(\Dnt\) be any matrix that is a left inverse of \(A_n\), i.e. such that \(\Dnt A_n = I_{n+3}\). Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the \(\Dnt\)'s, we obtain a lower triangular system \citep{dunkl2014orthogonal}, which can be expanded to obtain the recurrence:
\begin{align}
\begin{cases}
\bigPab_{-1}(x,y) := 0 \\
\bigPab_{0}(x,y) := P^{(a,b)}_{0,0} \\
\bigPab_{n+1}(x,y) = -\Dnt (B_n-G_n(x,y)) \bigPab_n - \Dnt C_n  \,\bigPab_{n-1}(x,y), \quad n = 1,2,3,\dots,N.
\end{cases}
\end{align}

@@

We can obtain the matrix operator for \(\Delta\), that will take us from coefficients for expansion in the weighted space
\[
\Wii(x,y) \: \Pii(x,y)
\]
to coefficients in the non-weighted space
\[
\Pii(x,y).
\]
This operator will have Banded-Block-Banded structure. We first note that, for general \(a,b\) and general function \(f(x,y) = \mathbf{f}^T \: \bigPab(x,y)\), there exist Banded-Block-Banded matrix operators \(D_x^{(a,b)}, \: D_y^{(a,b)}, \: W_x^{(a,b)}, \: W_y^{(a,b)}\) such that:
\begin{align}
\dx[f(x,y)] &= \bigP^{(a+1,b+1)}(x,y)^T \: D_x^{(a,b)} \: \mathbf{f}, \\
\dy[f(x,y)] &= \bigP^{(a,b+1)}(x,y)^T \: D_y^{(a,b)} \: \mathbf{f}, \\
\dx[\Wab(x,y) \: f(x,y)] &= W^{(a-1,b-1)}(x,y) \: \bigP^{(a-1,b-1)}(x,y)^T \: W_x^{(a,b)} \: \mathbf{f}, \\
\dy[\Wab(x,y) \: f(x,y)] &= W^{(a,b-1)}(x,y) \: \bigP^{(a,b-1)}(x,y)^T \: W_y^{(a,b)} \: \mathbf{f},
\end{align}
Further, we note that there exist Banded-Block-Banded matrix operators \(T^{(1,0)\to(0,0)}\) and \(T^{(0,1)\to(1,1)}\) such that:
\begin{align}
x \bigP^{(1,0)}(x,y) &= \Big(T^{(1,0)\to(0,0)} \Big)^T \: \bigP^{(0,0)}(x,y) \\
\bigP^{(0,1)}(x,y) &= \Big(T^{(0,1)\to(1,1)} \Big)^T \: \bigP^{(1,1)}(x,y)
\end{align}
Then, the operator \(\Delta\) is given by
\begin{align}
    \Delta := D_x^{(0,0)} \: W_x^{(1,1)} + T^{(0,1)\to(1,1)} \: D_y^{(0,0)} \: T^{(1,0)\to(0,0)} \: W_y^{(1,1)}
\end{align}

@@

The goal here is to eventually investigate sparse spectral methods on triangulations of the sphere as an alternative to the spherical harmonics approach currently in use at the European Centre for Medium-range Weather Forecasts (ECMWF) in their weather and climate model. The aim would be to preserve the predictive skill of the current approach whilst avoiding the parallel scalability bottleneck from the global spectral transform, which is expected to inhibit the future performance of the ECMWF model.

>>>>

The goal of my research is to eventually investigate sparse spectral methods on triangulations of the sphere as an alternative to the spherical harmonics approach currently in use at the European Centre for Medium-range Weather Forecasts (ECMWF) in their weather and climate model. The aim would be to preserve the predictive skill of the current approach whilst avoiding the parallel scalability bottleneck from the global spectral transform, which is expected to inhibit the future performance of the ECMWF model.

This document is to provide a brief outline of my work to date, with a closer look at our most recent area of research (solving PDEs on the half disk using sparse spectral methods). The work outlined in this report serves as initial steps to pave the way to solving PDEs on spherical triangles. All code used thus far and from hence forth will mainly be written in the Julia programming language.

@@

We can the define the spherical harmonics, orthogonal on the unit sphere as \cite{DLMFSphericalandSpheroidalHarmonics, DLMFFerrers, DLMFAssociatedLegendre}:
\begin{align}
\Ylmfull &:= \clm \eimphi \Plm (\costheta) \\
&= \alphalm (1 - (\costheta)^2)^\frac{|m|}{2} \eimphi P^{(|m|,|m|)}_{l-|m|}(\costheta), \quad \text{where } 0 \le |m| \le l, \, l \in \No.
\end{align} 
Note that we can then express \(\Ylm\) in terms of \(x,y,z\) instead of \(\theta, \varphi\) by noting that \(\costheta = z\) and that \( \eimphi\) can be expressed in terms of \(x,y,z\) for any \(m\in\mathbb{Z}\). Indeed, they are polynomials in $x,y,z$ which we denote $\Ylm(x,y,z)$. They span all polynomials modulo the ideal generated by $x^2 + y^2 + z^2 - 1$. While this is equivalent to standard techniques in the litrature concerning spherical harmonics, we approach this with our goal in mind of developing sparse spectral methods on spherical triangles. Thus, we try to use language and formulations that will translate as we move through our steps towards that goal.

\citep{dunkl2014orthogonal} \bstodo{Try to always include page numbers in citations to books, as otherwise you'll forget where it is} provided a framework for gaining a recurrence relation for each \(\bigP_{l+1}\) in terms of the previous two sub-vectors \(\Pl\) and \(\bigP_{l-1}\), utilising the sub matrices that make up the block-tridiagnonal Jacobi operator matrices for multiplication by $x, y$ and $z$. Clenshaw's algorithm then gives us a way of evaluating a function given its expansion in the spherical harmonic basis (up to a finite order). This provides an optimal complexity method for evaluating spherical harmonic expansions. This is an alternative to using the explicit expressions that may have benefits, though further investigation is required. \bstodoinline{Reword above}

@@

For \(l \in \No\), \(m \in \Z\) s.t. \(0 \le |m| \le l\):
\begin{align}
x\,\Ylm(x,y,z) &= \Alm Y^{m+1}_{l+1}(x,y,z) +  \Blm Y^{m+1}_{l-1}(x,y,z) \nonumber \\
		     & \quad \quad \quad + \Dlm Y^{m-1}_{l+1}(x,y,z) + \Elm Y^{m-1}_{l-1}(x,y,z), \\
y\,\Ylm(x,y,z) &= - i \, \big[\Alm Y^{m+1}_{l+1}(x,y,z) +  \Blm Y^{m+1}_{l-1}(x,y,z) \big] \nonumber \\
		     &\quad \quad \quad + i \, \big[ \Dlm Y^{m-1}_{l+1}(x,y,z) + \Elm Y^{m-1}_{l-1}(x,y,z) \big], \\
z\,\Ylm(x,y,z) &= \Flm Y^{m}_{l+1}(x,y,z) + \Glm Y^{m}_{l-1}(x,y,z) ,
\end{align}
where \(\Alm, \Blm, \Dlm, \Elm, \Flm, \Glm \in \R\) have explicit derivable and calculable formulae.

@@

For each \(l \in \No\) let \(\Dlt\) be any matrix that is a left inverse of \(A_l\), i.e. such that \(\Dlt A_l = I_{2l+3}\). Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the \(\Dlt\)'s, we obtain a lower triangular system \citep{dunkl2014orthogonal} \bstodo{Try to always include page numbers in citations to books, as otherwise you'll forget where it is}, which can be expanded to obtain the recurrence:
\begin{align}
\begin{cases}
\bigP_{-1} := 0 \\
\bigP_{0} := \alpha^0_0 \\
\bigP_{l+1} = -\Dlt (B_l-G_l) \Pl - \Dlt C_l  \,\bigP_{l-1}, \quad l \in \N.
\end{cases}
\end{align}

\subsection{Evaluation of a scalar function on the sphere}
We can use a multivariate analogue of the Clenshaw algorithm to evaluate a function at a given point \((x,y,z)\) on the unit sphere provided we know the coefficients of the function when expanded in the spherical harmonic basis, i.e. suppose \(f(x,y,z)\) is a function and we know the set \(\{\bold{f}_l\}\) s.t.
\begin{align}
f(x,y,z) \approx \sum^N_{l=0} \bold{f}_l^T \, \Pl (x,y,z), \quad \text{where } \Pl (x,y,z), \bold{f}_l \in \R^{2l+1} \text{ for each } l \in \{0,\dots,N\}.
\end{align}

@@

By noting that the tangent space is then spanned by $\{\gradYlm, \gradpYlm\}$, and that these gradients and perpendicular gradients are orthogonal polynomials themselves, we can use the same ideas to create a framework for functions in the tangent space as expansions in this OP basis. Note that the perpendicular gradient is related to the regular surface gradient by
\begin{align}
\gradpYlm(x,y,z) = \unitvec \times \gradYlm(x,y,z),
\end{align}
where \(\unitvec\) is the unit vector normal to the surface of the sphere at the point \((x,y,z)\), i.e. as we are looking at the unit sphere, \(\unitvec\) is simply given by
\begin{align}
\unitvec = \begin{bmatrix} x \\ y \\ z \end{bmatrix}.
\end{align}
For any scalar spherical harmonic (SSH) $\Ylm$, the triad of $\gradYlm, \: \unitvec \times \gradYlm, \: \unitvec \Ylm$ can be referred to as vector spherical harmonics \cite{BarreraEtAl}.

Since the spherical harmonics are a basis for the surface of the sphere, and the tangent space of the sphere is spanned by the gradient  and perpendicular gradient of a scalar function, we have that the gradients and perpendicular gradients of the spherical harmonics are a basis for the tangent space, namely \(\gradYlm\), \(\gradpYlm\).

@@

The gradient and perpendicular gradient of a spherical harmonic \(\Ylm(x,y,z)\) can be expressed in terms of spin-1 tensor spherical harmonics, which in turn can each be expressed as a vector-weighted sum of spherical harmonics.

@@

Then using the operators on the coefficient vectors, the timestepping method is: 
\begin{align}
\mathbf{u}^c_{n+1} &= \mathbf{u}^c_{n} + \Delta t \: (G \mathbf{h}^c_{n+1} - F K \mathbf{u}^c_{n+1}) \\
\mathbf{h}^c_{n+1} &= \mathbf{h}^c_n - \Delta t \: \mathcal{H} D \mathbf{u}^c_{n+1}
\end{align}
$\iff$
\begin{align}
\mathbf{u}^c_{n+1} &= (I + \Delta t F K + \Delta t^2 \mathcal{H} G D)^{-1} \: (\mathbf{u}^c_{n} + \Delta t \: G \mathbf{h}^c_n) \\
\mathbf{h}^c_{n+1} &= \mathbf{h}^c_n - \Delta t \: \mathcal{H} D \mathbf{u}^c_{n+1}
\end{align}
where $F, K, G, D$ are all operators corresponding to multiplying by the Coriolis parameter, outward normal vector cross product, gradient and divergence operations respectively. It is important to note that these can be easily constructed due to the relationships between the vector spherical harmonics and scalar spherical harmonics we have outlined, and have \bstodo{Banded-Block-Banded?} a sparse Block-Banded structure.

@@

As a stepping stone to working on spherical triangles, we will try to construct orthogonal polynomials on a circle arc, given by $x \in [h, 1]$, $y = (1-x^2)^\frac{1}{2}$. We will use the Lanczos algorithm as implemented in the ApproxFun package \cite{ApproxFun} to obtain 1D OPs $\{T^h_n\}$ (orthogonal wrt $w(x) = (1-x^2)^{-\frac{1}{2}}$) and $\{U^h_n\}$ (orthogonal wrt $w(x) = (1-x^2)^{\frac{1}{2}}$) on the interval $[h,1]$, and create our 2D OPs according to:\sotodo{Cite Huybrechs reference for connection in literature}
\begin{align}
\bigP_0(x,y) := T^h_0(x) \equiv T_0,
\quad \quad 
\bigP_n(x,y) := \begin{bmatrix}
		T^h_n(x) \\
		y \: U^h_{n-1}(x)
	\end{bmatrix} \in \R^2 \quad \forall n \in \N,
\quad \quad 
\bigP := \begin{bmatrix}
		\bigP_0 \\
		\hline
		\bigP_1 \\
		\hline
		\bigP_2 \\
		\vdots \\
	\end{bmatrix}.
\end{align}

@@

By using a similar process to \cite{dunkl2014orthogonal} (p??) \bstodo{Try to always include page numbers in citations to books, as otherwise you'll forget where it is} we can construct 2D orthogonal polynomials on the half disk from 1D orthogonal polynomials on the intervals \([0,1]\) and \([-1,1]\). We denote the half disk by \(\Omega\) where
\begin{align}
\Omega := \{(x,y) \in \R^2 \quad | \quad 0 < x < 1, \: -(1-x^2)^{\half} < y < (1-x^2)^{\half}\}.
\end{align}

@@

Fix \(a,b \in \R\). Then for any function \(f : \Omega \to \R\) we can approximate (or exactly express) \(f\) by
\begin{align}
f(x,y) \approx \sum_{n=0}^N \bigP_n^{(a,b)}(x,y)^T \: \mathbf{f}_n
\end{align}
where
\begin{align}
\bigP(x,y) := \begin{bmatrix}
		\bigP_0(x,y) \\
		\hline
		\bigP_1(x,y) \\
		\hline
		\bigP_2(x,y) \\
		\hline
		\vdots \\
		\hline
		\bigP_N(x,y)
	\end{bmatrix},
\quad
\bigP_n(x,y) := \begin{bmatrix}
		P_{n,0}(x,y) \\
		\vdots \\
		P_{n,n}(x,y)
	\end{bmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N,
\end{align}
and where
\begin{align}
\mathbf{f} &:= \begin{bmatrix}
		\mathbf{f}_0 \\
		\hline
		\mathbf{f}_1 \\
		\hline
		\mathbf{f}_2\\
		\hline
		\vdots \\
		\hline
		\mathbf{f}_N
	\end{bmatrix},
\quad
\mathbf{f}_n := \begin{bmatrix}
		f_{n,0} \\
		\vdots \\
		f_{n,n}
	\end{bmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N, \\
f_{n,k} &:= \frac{\langle f, \: \Pnk^{(a,b)} \rangle_{W^{(a,b)}}}{|| \Pnk^{(a,b)} ||_{W^{(a,b)}}}
\end{align}

@@

For each \(n = 0,1,\dots,N-1\) let \(\Dnt\) be any matrix that is a left inverse of \(A_n\), i.e. such that \(\Dnt A_n = I_{n+3}\). Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the \(\Dnt\)'s, we obtain a lower triangular system \citep{dunkl2014orthogonal} \bstodo{Try to always include page numbers in citations to books, as otherwise you'll forget where it is}, which can be expanded to obtain the recurrence:
\begin{align}
\begin{cases}
\bigPab_{-1}(x,y) := 0 \\
\bigPab_{0}(x,y) := P^{(a,b)}_{0,0} \\
\bigPab_{n+1}(x,y) = -\Dnt (B_n-G_n(x,y)) \bigPab_n - \Dnt C_n  \,\bigPab_{n-1}(x,y), \quad n = 1,2,3,\dots,N.
\end{cases}
\end{align}

@@

We can obtain the matrix operator for \(\Delta\), that will take us from coefficients for expansion in the weighted space
\[
\Wii(x,y) \: \bigP^{(1,1)}(x,y)
\]
to coefficients in the non-weighted space
\[
\bigP^{(1,1)}(x,y).
\]
Importantly, this operator will have Banded-Block-Banded structure, and hence will be sparse. Note that this construction will ensure the imposition of the Dirichlet zero boundary conditions on $\Omega$.

For general \(a,b\) and general function \(f(x,y) = \mathbf{f}^T \: \bigPab(x,y)\), there exist Banded-Block-Banded matrix operators \(D_x^{(a,b)}, \: D_y^{(a,b)}, \: W_x^{(a,b)}, \: W_y^{(a,b)}\) such that:
\begin{align}
\dx[f(x,y)] &= \bigP^{(a+1,b+1)}(x,y)^T \: D_x^{(a,b)} \: \mathbf{f}, \\
\dy[f(x,y)] &= \bigP^{(a,b+1)}(x,y)^T \: D_y^{(a,b)} \: \mathbf{f}, \\
\dx[\Wab(x,y) \: f(x,y)] &= W^{(a-1,b-1)}(x,y) \: \bigP^{(a-1,b-1)}(x,y)^T \: W_x^{(a,b)} \: \mathbf{f}, \\
\dy[\Wab(x,y) \: f(x,y)] &= W^{(a,b-1)}(x,y) \: \bigP^{(a,b-1)}(x,y)^T \: W_y^{(a,b)} \: \mathbf{f}.
\end{align}
The incrementing and decrementing of parameters as seen here is analogous to other well known orthogonal polynomials families' derivatives, for example the Jacobi polynomials, as seen in the DLMF \cite{DLMFDerivatives}.
Further, we note that there exist Banded-Block-Banded matrix operators \(T^{(1,0)\to(0,0)}\) and \(T^{(0,1)\to(1,1)}\) such that:
\begin{align}
x \bigP^{(1,0)}(x,y) &= \Big(T^{(1,0)\to(0,0)} \Big)^T \: \bigP^{(0,0)}(x,y) \\
\bigP^{(0,1)}(x,y) &= \Big(T^{(0,1)\to(1,1)} \Big)^T \: \bigP^{(1,1)}(x,y)
\end{align}
Then, the operator \(\Delta\) is given by
\begin{align}
    \Delta := D_x^{(0,0)} \: W_x^{(1,1)} + T^{(0,1)\to(1,1)} \: D_y^{(0,0)} \: T^{(1,0)\to(0,0)} \: W_y^{(1,1)}
\end{align}

@@

Beyond this, solving PDEs on isosceles triangles should follow using the same procedure. However, we are expecting that to solve PDEs on spherical triangles (as is our ultimate goal) will require further ideas beyond the procedures given in this document.

\sotodoinline{Expand bibliography: cite spherical harmonics books, Huybrechs, ECMWF paper, etc.}
